Parsing Postfix log files

### 2007-10-16 doc/logparser.tex
Greatly improved the explanation of the dips in the normal/shuffle/reverse factor graphs.  I'm quite pleased with it, though the paragraph pulling it all together might need work.

The resulting execution times are plotted in graph~\ref{normal vs shuffled
vs reversed ordering}, where the difference is not readily apparent, but
graphs~\ref{normal vs shuffled vs reversed ordering factor} show the
percentage increase of execution times (mean and standard deviation are
shown after the graph).  Overall this provides a modest but worthwhile
performance increase.

@@

\subsubsection{Normal vs shuffled vs reversed ordering}
\label{normal vs shuffled vs reversed ordering}
\includegraphics{build/plot-normal-shuffle-reverse.ps}

@@

The dips at logfiles 22 and 62--68 correspond to peaks in log file size in
graph~\ref{execution time vs file size vs number lines}.  The explanation
for this took some time to arrive at, but it turns out to be reasonably
simple.  Examining the database after parsing logfile~22 in isolation shows
64 rules matched during parsing, but as shown in graph~\ref{log22
distribution by rule} the distribution of matching rules is extremely
skewed.  Comparing the number of log lines per program versus rules per
program in graphs~\ref{log22 log line and rule distribution by program}
shows that \texttt{postfix/qmgr} has the highest number of log lines,
however it has a low number of rules.  This combination results in a low
average number of rules to try for \texttt{postfix/qmgr} log lines,
reducing the overall average number of rules to try per log line.  In turn
this reduces the average parsing time per log line, and thus the parsing
time for the log file.  While the parsing time is substantially greater
than the time for the log files on either side, the increase from optimal
ordering to reversed ordering is proportionally smaller, producing the dips
in graph~\ref{normal vs shuffled vs reversed ordering factor}.

This non-intuitive effect will not be exhibited by log files in general: in
a normally functioning mail system the impact of rule ordering will scale
linearly with log file size.  The skewed distribution of log lines in the
specific log files creates the anomaly described herein.

@@

\subsubsection{Log 22 distribution}
\label{log22 distribution by rule}
\includegraphics{build/plot-log22-distribution-by-rule.ps}

The individual rules are not labelled on the x-axis because doing so makes
the graph illegible.

\subsubsection{Log 22 log line and rule distribution by program}
\label{log22 log line and rule distribution by program}

\noindent
\includegraphics{build/plot-log22-rules.ps}

\noindent
\includegraphics{build/plot-log22-log-lines.ps}

As shown in graphs above, the most common log lines are from programs which
have a low number of rules, thus reducing the average parsing time per
line, and the overall execution time relative to the input size.

\subsubsection{Rule hits}
\label{rule hits}

@@

>>>>

Graphs~\ref{normal vs shuffled vs reversed ordering factor} show the
percentage increase of execution times (mean and standard deviation are
shown after the graph).  Overall this provides a modest but worthwhile
performance increase.

@@

@@

\subsubsection{Why are there dips in the graphs?}
\label{Why are there dips in the graphs?}

The dips at logfiles 22 and 62--68 correspond to peaks in log file size in
graph~\ref{execution time vs file size vs number lines}, and peaks in
graph~\ref{execution time vs file size vs number lines factor} (where a
peak means that more lines are processed per second, i.e.\ performance is
better).  The explanation for this took some time to arrive at, but it
turns out to be reasonably simple.  The large log files in question were
caused by a mail forwarding loop, where the distribution of log lines is
quite different to normal, resulting in different performance
characteristics.

The mail loop was set up by a user modifying his mail forwarding to:
\newline \hspace*{2em}\texttt{$\backslash$username, username@domain}
\newline This instructs Postfix to deliver the mail to the local user, and
also forward to the remote address; this is generally not a problem except
that the remote address in this case is the user's address, creating an
infinite loop.  To prevent this happening Postfix examines the Delivered-To
header in the mail and if the mail has already been delivered to the
current address it is bounced back to the sender with the error message
\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
well, but unfortunately the user noticed they had not received any mail in
a while, and opted to send a test mail to themselves, causing a loop not
caught by Postfix as described below.

    \item Postfix accepts a mail from username@domain, for
        username@domain.

    \item Postfix delivers the mail to the local mailbox and
        username@domain, as instructed by the user's forwarding
        instructions. The forwarded mail has a
        \texttt{Delivered-To:~username@domain} header added, and the
        envelope sender address is username@domain.  Log lines are added by
        \textit{postfix/local}, \textit{postfix/qmgr\/} (twice),
        \textit{postfix/cleanup\/} and finally \textit{postfix/pickup}.

    \item Postfix accepts the mail for username@domain, but while
        delivering it notices that the \texttt{Delivered-To} header already
        contains the address it's currently delivering to, and therefore
        sends a bounce notification with sender address \textit{$<>$\/} to
        the original sender: username@domain.  Log lines are added by
        \textit{postfix/local}, \textit{postfix/qmgr\/} (twice) and
        \textit{postfix/cleanup}.

    \item Postfix delivers the bounce notification to both the local
        mailbox and to username@domain, as instructed by the user's
        forwarding instructions.  A \texttt{Delivered-To:~username@domain}
        header is added to the forwarded bounce notification, and the
        envelope sender address is now username@domain.  Log lines are
        added by \textit{postfix/local},
        \textit{postfix/qmgr\/} (twice), \textit{postfix/cleanup}, and finally
        \textit{postfix/pickup}.

    \item Postfix accepts the forwarded bounce notification but while
        delivering it notices that the \texttt{Delivered-To} header already
        contains the address currently being delivered to, and sends a
        bounce notification to the sender: username@domain.  Log lines are
        added by \textit{postfix/local}, \textit{postfix/qmgr\/} (twice)
        and \textit{postfix/cleanup}.

    \item GOTO 2.

The sequence described above occurs extremely rapidly, as Postfix does not
have to deliver the mail to an external system, so mails are delivered,
bounced and generated as fast as the disks can keep up, resulting in a huge
volume of logs.

The vast majority of log lines when a mail loop occurs are from Postfix
components which have a small number of rules associated with them, whereas
in general the most common log lines are logged by \textit{postfix/smtpd},
which also has the highest number of rules.  \textit{postfix/smtpd\/} log
lines are distributed across rules much more evenly that
\textit{postfix/qmgr}, \textit{postfix/local}, \textit{postfix/cleanup\/}
or \textit{postfix/pickup}, so the average number of rules required to
parse a \textit{postfix/smtpd\/} log line is much higher that the average
number required to parse other log lines.  Thus the average number of rules
required to parse a random line from logs containing a mail loop is much
lower than the average number required to parse a random log line from logs
without a mail loop, as can be seen in the peaks in graphs~\ref{execution
time vs file size vs number lines factor} and~\ref{normal vs shuffled vs
reversed ordering factor}.  The difference is more extreme in the reversed
ordering case because the largest number of rules to be consulted for the
commonly occurring log lines during a mail loop is far less than the
largest number of rules that are consulted for commonly occurring lines
outside of a mail loop.  The number of rules which needs to be consulted
when the ordering is shuffled varies between the optimum and nadir, and
the performance varies proportionally.

The difference between logs with a mail loop and logs without can be seen
in this table showing the increases for the different rule orderings and
combinations of log files:

@@

@@

\subsection{Rule hits}
\label{rule hits}
